### 2020-06-12 Manuscript/manuscript.tex
remove Cuyania text per Athena's suggestion; add Paleo-Iapetus hypothesis as explanation of Laurentia's eastern margin rift history; add Amazonia section to the in-development paleogeographic neighbors section

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.

The tectonic history of terrane transfer, such as Cuyania, can provide valuable paleogeographic constraints. Cuyania is widely interpreted be a rifted fragment of southeast Laurentia that separated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton, as the record that lingers on the craton is similar. One interpretation is that there was successful break-up along the eastern margin during the ca. 580 to 550 Ma interval of rifting prior to the ca. 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent. The Maz-Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting. Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca. 530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca. 460 Ma Famatinian orogeny in West Gondwana \citep{Rapalini2018a}.

@@

@@

%\subsection{Evaluating Laurentia's paleogeographic neighbors}
%%
%%\subsubsection{Baltica}
%%
%\subsubsection{North China}
%
%The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP (\citealp{Zhao2019a}; Zhang et al., this volume). While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. \cite{Zhao2019a} also argued that similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival of North China is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the ca. 1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and Zhang et al. (this volume) to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. In terms of departing from this position, one possibility is that

>>>>

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar. Recognizing this ambiguity, \citet{Robert2020a} propose that rather than being associated with spatially-restricted or failed rifting that ca. 700 Ma extension is associated with breakup and separation of Laurentia and its conjugate continent (that they interpret to be Amazonia). This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Thethys). Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia ca. 600 Ma leading to the formation of the Neo-Iapetus Ocean and the record of passive margin development on Laurentia \citep{Robert2020a}.

@@

\subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
\textit{Note that this section is currently under development and is incomplete}

\subsubsection{Amazonia}

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of Amazonia lithosphere that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest the Suns\'as orogen of Amazonia as the best match for southern and central Appalachian inliers. This positioning leads to a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Evans2013a,Cawood2017a}. In this model, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a}.

%%\subsubsection{Baltica}
%%
\subsubsection{North China}

The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP (\citealp{Zhao2019a}; Zhang et al., this volume). While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. \cite{Zhao2019a} also argued that similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the ca. 1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and Zhang et al. (this volume) to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. In terms of departing from this position, one possibility is that its departure is associated with early Neoproterozoic extension in northwest Laurentia.

@@
